
\documentclass[paper=a4,12pt,DIV16,BCOR8mm,fleqn,twoside]{scrartcl}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[english]{babel}

\usepackage[pdftex]{graphicx}
\usepackage{xspace}
\usepackage{textcomp}

% needed for FloatBarrier
\usepackage[section]{placeins}

\usepackage{verbatim}
\usepackage[nomargin,inline,draft]{fixme}

\usepackage{comment}

%\usepackage[firstpage]{draftwatermark}
%\SetWatermarkScale{4}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{txfonts}


\usepackage{booktabs}
\usepackage{rotating}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{multicol}

\newcommand{\sturz}[2]{\rotatebox{90}{\parbox{#2}{\center #1}}}

\usepackage{listings}
\input{listing}
\input{newcommands}


\usepackage{float}
 
%\usepackage{chngcntr}
%\counterwithout{footnote}{chapter}

% need to appear as last \usepackage to avoid nonsensical warnings,
% cf. http://mrunix.de/forums/showthread.php?t=38657#8
\usepackage[pdftex]{hyperref}

\floatstyle{plain}
%\newfloat{listing}{thp}{lop1}[chapter]
\floatname{listing}{Listing}

%\newfloat{predfun}{thp}{lop2}[chapter]
\floatname{predfun}{Listing}


% share counters of all floats:
\makeatletter
\let\c@listing\c@figure
\let\c@predfun\c@figure
%\let\c@lemma\c@figure
\let\c@table\c@figure
\makeatother

% suppress hyperref warnings
\makeatletter
\def\Hy@Warning#1{}
\def\Hy@WarningNoLine#1{}
\makeatother

\renewcommand\floatpagefraction{.90}
\renewcommand\topfraction{.90}
\renewcommand\bottomfraction{.90}
\renewcommand\textfraction{.1}
\setlength{\unitlength}{1mm}

\setlength\topsep{1.5\baselineskip}  % larger space around theorem-like environments

\addtolength{\parskip}{0.5\baselineskip}
\parindent=0cm

\setcounter{tocdepth}{1}
\setcounter{secnumdepth}{2}

\begin{document}

\title{Specification Concepts for Packages of Subset 026 (Chapter 7)}
\author{Jens Gerlach}
\maketitle

\section{Predicates and Logic Functions of \inl{BitStream}}

The type \inl{Bitstream} models a bit stream and is given
by a byte array of \inl{size} elements that starts at \inl{addr}.
A bit stream also holds a \emph{bit position} that marks
the current position in the bit stream from where bits can be
read or to which bits can be written.

\begin{lstlisting}[style = acsl-block]
struct Bitstream
{
    uint8_t*  addr;
    uint32_t  size;
    uint32_t  bitpos;
};

typedef struct Bitstream Bitstream;
\end{lstlisting}

\begin{itemize}
\item
\inl{predicate Readable{L}(Bitstream* stream)}

This predicate is true if all bytes in the range
\inl{[stream->addr[0..size-1]} are valid for reading.

\item
\inl{predicate Writeable{L}(Bitstream* stream)}

This predicate is true if all bytes in the range
\inl{[stream->addr[0..size-1]} are valid for reading \emph{and} writing.

\item
\inl{predicate Invariant{L}(Bitstream* stream, integer length)}

This predicate is equivalent to the following condition

\begin{lstlisting}[style = acsl-block]
   8 * stream->size        <= UINT32_MAX   &&
   stream->bitpos + length <= UINT32_MAX;
\end{lstlisting}

\item
\inl{predicate Normal{L}(Bitstream* stream, integer length)}

This predicate is true if \inl{stream->bitpos + length <= 8 * stream->},
that is if at least \inl{length} bits can be read from or written to the
bitstream.

\end{itemize}

\section{Predicates and Logic Functions of Packages}


Let \inl{P} be the type of a package.

\begin{lstlisting}[style = acsl-block]
   struct P {
      ...
   };

   typedef struct P P;
\end{lstlisting}

We will use follwing \acsl predicates and logic functions with the following signatures
for the specification of decoder and encoder functions for \inl{P}.

\begin{itemize}
\item
\inl{logic integer MaxBitSize{L}(P* p)}

Returns the number of bits that might be read/written by \inl{P}.

\item
\inl{logic integer BitSize{L}(P* p)}

Returns the number of bits that have been read/written by \inl{P}.
This number is always less or equal than \inl{MaxBitSize}.

\item
\inl{predicate Invariant(P* p)}

This predicate is true if all type invariants of \inl{P}
are satisified.

\item
\inl{predicate ZeroInitialized(P* p)}

This predicate is true is all relevant elements of \inl{p} are
equal to zero.

\item
\inl{predicate EqualBits(Bitstream* stream, integer pos, P* p)}

This predicate is true if starting from the bit position
\inl{pos} in the bit stream \inl{stream} the following
\inl{BitSize} number of bits are equal to the corresponding bits
in the elements of \inl{p}.

\inl{predicate UpperBitsNotSet(P* p)}

This predicate is true if the appropriate upper (higher) bits
in the elements of \inl{p} are \emph{not} set.

\item
\inl{Separated(Bitstream* stream, P* p)}

This predicate is true if there is now aliasing between the elements of
\inl{p} and those of \inl{Bitstream}.

\end{itemize}

%\listoffixmes
%\tableofcontents
%\listof{predfun}{List of Logic Specifications}
%\listoffigures
%\listoftables


% jg:  the "phantomsection" is a hack to fix the wrong reference to bibliography
% see also http://sumanta679.wordpress.com/2009/05/15/latex-list-of-table-list-of-figures-and-bibliography-in-toc/
%\newpage
%\phantomsection \label{listoffig}
%\addcontentsline{toc}{chapter}{Bibliography}
%\bibliographystyle{unsrt}
%\bibliography{bibliography}

\end{document}

